\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2017-10-24 document/XSGRB_sample.tex
Adding ontro

\usepackage{url}
\usepackage{color,hyperref}
\definecolor{linkcolor}{rgb}{0,0.3,0.7}
\hypersetup{colorlinks=true,
	linkcolor=linkcolor, 
	citecolor=linkcolor,
	filecolor=linkcolor, 
	urlcolor=linkcolor}

@@

\usepackage{xcolor}
\newcommand\todo[1]{\textbf{(#1)}}

@@

\title{XSGRB - The X-shooter GRB afterglow sample
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}}

\titlerunning{The X-shooter sample of GRBs}

\author{
	J.~Selsing\inst{1}, 
	D.~Malesani\inst{1}, 
	T.~Kr\"{u}hler\inst{1}, 
	J.~P.~U. Fynbo\inst{1}, 
	P.~Goldoni\inst{14}, 
	A.~de~Ugarte~Postigo\inst{11}, 
	J.~Japelj\inst{20},
	P.~D'Avanzo,
	Z.~Cano,
	S.~Covino\inst{10}, 
	V.~D'Elia\inst{7, 12}, 
	H.~Flores,
	O.~E.~Hartoog\inst{6},
	J.~Hjorth\inst{1}, 
	P.~Jakobsson\inst{5}, 
	A.~Levan,
	A.~Melandri,
	S.~Piranomonte \inst{7},
	R.~S\'anchez-Ram\'irez\inst{11},
	S.~Schulze\inst{17, 18}, 
	N.~R.~Tanvir\inst{19},
	C.~Th{\"o}ne,
	S.~D.~Vergani\inst{7, 8},
	P.~M.~Vreeswijk\inst{3}, 
	D.~J.~Watson\inst{1},
	K.~Wiersema\inst{19},
	D.~Xu\inst{1}
	L.~Christensen\inst{1},
	A.~De~Cia\inst{3}, 
	L.~Kaper\inst{6}, 
	L.~A.~Antonelli,
	F.~Fiore,
	A.~Gomboc,
	P.~Groot,
	F.~Hammer,
	C.~Ledoux\inst{2}, 
	E.~Maiorano,
	B.~Milvang-Jensen\inst{1}, 
	E.~Palazzi,
	E.~Pian,
	J.~Schaye,
	G.~Tagliaferri\inst{7},
	R.~A.~M.~J.~Wijers\inst{6}
}

@@

\abstract{
In this work we present the spectra of \textit{all} $\gamma$-ray bursts (GRBs)
afterglows that have \textit{ever} been observed at the VLT/X-shooter. This
totals 137 spectroscopic observation of 118 individual bursts, of which
redshifts have been secured for 91 \% covering a redshift range from 0.059 to
$\sim$ 8.5. Based on a set of unbiasing selection criteria, the follow-up effort
has been guided by the \textit{Swift}-telescope to produce a homogeneous sample
of afterglows for 86 spectra, representing the underlying \textit{Swift} GRB population.
We here provide a public release of all the spectra, including continuum
estimates and telluric corrections. We provide an assessment of the degree of
completeness with respect to the parent GRB population, in terms of the X-ray
properties of the bursts in the sample and find that the sample presented here
can not be significantly distinguished from the \textit{Swift} sample.
For all bursts (39) for which it is possible, we provide a measurement of the
optically derived hydrogen column density, increasing the total number of
published HI column densities by $\sim$ 50 \% and characterize the distribution
compared with literature values.
We constrain the fraction of dark burst to be 27\% - 42\% and confirm previous
results that higher optical darkness has increased X-ray excess
absorption.
%This dataset provides a unique possibility to study the ISM across cosmic time.
}

@@

For a brief moment, the energy output of a $\gamma$-ray burst (GRB) can rival
that of the stellar light integrated over the entirety of the observable
universe \citep{Kumar2015}. Bright across the entire electromagnetic spectrum,
the following afterglow makes possible the exploration of the intervening space,
otherwise dark. Since the discovery of the first $\gamma$-ray counterparts
nearly two decades ago\citep{Costa1997a, VanParadijs1997, Djorgovski1997,
	Frail1997}, significant advances in the understanding of the underlying
phenomenology and potential usage has been made.

Categorized on the basis of their duration and spectral hardness, two main classes of GRBs exist, long and short \citep{Kouveliotou1993}. Now thoroughly established, the connection between long GRBs and supernovae (SNe) \citep{Galama1998, Hjorth2003a, Stanek2003, Woosley2006c} and the association of the location of long GRBs with the most star-forming regions in their host galaxies \citep{Fruchter2006, Svensson2010a}, make a stellar origin a likely explanation for the progenitor system. The seeming metal aversion \citep{Graham2013, Kruhler2015, Perley2016b} in selection of host at low redshift along with theoretical single-progenitor models preferring more pristine starting conditions \citep{Woosley2006d}, paints a picture where the core of a rapidly spinning, massive star undergoes gravitational collapse, although a certain degree of binarity \citep{Fryer2005, Podsiadlowski2010} in the progenitor population is required to explain the host metallicity distribution to super solar values \citep{Niino2011}, see \citet{Levan2016} for recent review. The driving engine in short GRBs has been established as the merger of two compact objects \citep{Metzger2010} through the detection of an associated kilonova \citep{Tanvir2013a, Berger2013a}. With the advent of the first gravitational wave (GW) detection \cite{Abbott2016a}, the electromagnetic counterpart in the form of a short

GRBs as cosmological probes

Do GRBs trace star formation

Afterglow spectroscopy

Complete samples

Content of paper

%Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
%towards about 10$^5$ QSOs have 5 systems with
%$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
%\citep{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
%systems in the majority of cases \citep{Jakobsson2006a, Fynbo2009}. Whereas DLAs
%towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
%atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
%\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
%star-forming galaxies over the full history of cosmic star formation from $z
%\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
%	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
%spectrum from X-rays to the sub-mm) we can hence characterize the properties of
%star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
%molecular contents, ISM temperatures, UV-flux densities, etc.. 
%
%
%This is, however,
%only possible as long as there are satellites in orbit that rapidly and
%accurately locate GRBs. The currently operating \textit{Swift} satellite,
%launched in 2004 and still fully-functioning, allows for very efficient
%follow-up observations of GRBs due to its unprecedented rate, speed and
%precision of localisations.
%
%There are several unanswered but fundamental questions that must be addressed in
%order to exploit the full potential of GRBs as cosmological probes. More than
%50\% of the \textit{Swift} bursts with measured redshift are at $z > 2$, and
%5--10\% are expected to be above $z=5$ \citep{Salvaterra2007,
%	Salvaterra2012,Jakobsson2012, Perley2016a}. A high redshift completeness is
%crucial for our understanding of the link between the number density of GRBs per
%unit redshift and the global star-formation history of the Universe, as measured
%by other means \citep[UV, FIR, sub-mm, see][]{Robertson2012}. The detection of
%GRBs at $z > 6$ shows that GRBs have become competitive as a tool to identifying
%galaxies at the highest redshifts and unsurpassed in providing detailed
%abundance information via absorption line spectroscopy \citep{Tanvir2012,
%	McGuire2016}.
%
%\smallskip
%
%%The sample of short GRBs with redshifts remains too small and too
%%incomplete to really
%%constrain their redshift distribution (Guetta \& Piran 2006, A\&A, 453, 823;
%%Salvaterra et al.\ 2008, MNRAS, 388, 6) making it of extreme importance to
%%enlarge as far as possible the number of known redshifts for this class of
%%sources.
%
%From March 2005 to March 2016 there have been about 350 {\it Swift} bursts
%complying with our sample selection criteria, and
%about half of them have measured redshifts. Among the latter subset, the team
%proposing these observations has measured about two thirds of the redshifts,
%mainly with FORS1/2 and X-shooter \citep[see][]{Fynbo2009, Jakobsson2012, Kruhler2012}. Our
%current aim is to build a sample superior to our previous low-resolution survey
%\citep{Fynbo2009}, both in terms of quantity and  quality of
%the spectra. 
%
%Due to
%the wide wavelength range of X-shooter with the same observation cover molecular
%H$_2$ absorption near the atmospheric cut-off and all the strong emission lines
%from the host in the NIR arm \citep[e.g.,][]{Friis2015}. In
%general, the wide wavelength coverage ensures that we always have features on
%which to base a redshift measurement as long as the afterglow is brighter than
%about 23 mag in either the $R$- or $z$-band. Frequently, emission lines are also
%detected from the underlying host, which also provide further information such
%as SFR and metallicity (the top right panel in Fig.~1 shows an example). Only
%for 7 out of more than 70 secured spectra could we not measure a redshift. With
%the X-shooter survey we provide {\bf metallicity measurements} for about 30\%
%(Voigt-profile fits) of the $z>1.7$ events. So far we have measured
%metallicities for more than 20 GRB afterglows with X-shooter. With the wide
%wavelength coverage of X-shooter we can study important chemical species as
%Zn\,\textsc{ii}, Cr\,\textsc{ii} and $\alpha$ elements over a much wider
%redshift range than what is possible with other instruments. As an example, we
%have measured a metallicity of $0.1 Z_\odot$ for GRB\,100219A at $z=4.669$
%(Th{\"o}ne et al.\ 2013, MNRAS, 428, 3590), $0.02Z_\odot$ for GRB\,111008A at
%$z=4.991$ (Sparre et al.\ 2014, ApJ, 785, 150) and $0.05Z_\odot$ for the
%$z=5.9125$ GRB\,130606A (Hartoog et al.\ 2015, A\&A, 580, 139). Reconciling the
%abundance patterns of GRB absorbers, other types of absorbers, QSO DLAs in
%particular, and old stars in the Local Group is an important long-term goal (see
%also Sparre et al.\ 2014, ApJ, 785, 150). Metallicities are also measured from
%host emission lines (Kr{\"u}hler et al.\ 2015, A\&A, 581, A125). GRB
%spectroscopy also allows us to determine the dust content of their environments,
%both through analysis of the depletion pattern and through measurement of the
%associated extinction (Japelj et al.\ 2015, A\&A, 451, 2050). This allows us to
%quantify the dust-to-metals ratio and its evolution with redshift (e.g., De Cia
%et al.\ 2013, A\&A, 560, 88; Zafar \& Watson 2013, A\&A, 560, 26). %Moreover,
%% GRB
%%afterglows allow investigation of random foreground absorbers, complementing
%%QSO sightline studies (e.g., Schulze et
%%al.\ 2012, A\&A, 546, 20).
%
%%In Fynbo et al. (2009, ApJS, 185,
%%526), we compare the host-galaxy HI
%%column density and metallicity (most of which has been measured by our group)
%%with a sample of damped \lya\ systems (DLAs) along QSO sightlines. We show that
%%most GRB-DLAs have a much larger column than QSO-DLAs.  The difference in
%%column density and metallicity for GRB and QSO sightlines shows that the total
%%cross-section for environments like those of GRBs must be less than a few
%%percent of the total cross-section for QSO-DLAs, and suggests that GRB
%%sightlines probe regions with high gas density.
%
%%The issue of production and {\bf release of Lyman continuum radiation} is
%%currently not fully understood, but the evidence suggests that star-forming
%%galaxies are dominating, in particular at the highest redshifts (e.g., Bianchi
%%et al.\ 2001, A\&A, 376, 1; Chen et al.\ 2007, ApJ, 667, L25). Also here
%%X-shooter spectroscopy will allow major progress due to the excellent blue
%%sensitivity.
%
%%GRB afterglows can probe dust in high-redshift star-forming galaxies
%%including exploring the {\bf 2175~\AA{} extinction bump}
%%(El\'iasd\'ottir et al.\ 2009, ApJ, 697, 1725; Zafar et al.\ 2011, A\&A,
%%732, 143).
%%Due to the broad wavelength  coverage we can do
%%this more accurately and over a broader range of redshifts with X-shooter.
%
%\smallskip
%
%We will also determine the frequency and properties of {\bf molecular
%	absorption} towards GRB absorbers. Molecular gas is a key element to catalyze
%the process of star formation, but prior to our program H$_2$ had been detected
%just in two cases (tentatively in Fynbo et al.\ 2006, A\&A, 451, L47; securely
%in Prochaska et al.\ 2009, ApJ, 691, L27). With our X-shooter program we have
%found three more systems (Kr\"uhler et al.\ 2013, A\&A, 557, 18; D'Elia et al.\
%2014, A\&A, 564, 38; Friis et al.\, 2015, MNRAS, 451, 167). We are currently
%analysing more of our spectra for less obvious molecular absorption and we
%expect to find more (a dedicated sample paper is addressing this issue).
%
%\vspace{0.2cm} A natural question to ask is: {\bf how long should this work
%	continue?} Our view is that we need to keep observing the afterglows as long as
%we have \textit{Swift} in operation. Also note that the program is still
%producing many papers and provides data for many theses (Box 9 and 10).
%\textit{Swift} is currently funded until 2018, but is likely to get more
%extensions given its overwhelming success. As mentioned above GRBs allow us to
%probe star-forming galaxies that are almost impossible to study in other ways
%both in terms of redshifts, galaxy luminosity function, and regions within
%galaxies. After 7 periods we have secured seven spectra of $z>4$ GRBs, of which
%three were of sufficient quality to allow abundance measurements (Th\"one et
%al.\ 2013, Sparre et al.\ 2014, Hartoog et al.\ 2015). GRBs offer the only way
%to derive chemical abundances for the gas phase of central, actively
%star-forming regions of high-$z$ galaxies. The program also maintains a very
%high discovery potential where we occasionally find something completely
%unexpected that provides interesting clues to puzzles in other fields, e.g.
%extinction of type Ia supernovae (Fynbo et al.\ 2014, A\&A, 572, 12). Each of
%these spectra are like precious jewels – it is a type of observation that can
%never be repeated and a class of sightlines that can only be studied while we
%have operating GRB satellites.
%
%It is also worth adding that we have build up a rather unique team spread over
%Europe from Granada to Reykjavik, which by now has reached a point where the
%distribution of night shifts, the scientific exploitation of the data is
%efficient and where we are open to all new members who wish to participate. As
%mentioned all data are public immediately.
%
%{\bf For all of these reasons, we need to keep building up the sample of GRB
%	afterglow spectra now as we may have to wait many years before a mission like
%	\textit{Swift} becomes available again.}
%
%A significant proportion of GRBs lack a bright optical afterglow (``{\bf dark
%	bursts}'', e.g., Jakobsson et al.\ 2004, ApJ, 617, L21; Melandri et al.\ 2012,
%MNRAS, 421, 1265). Some of these are at the highest redshifts ($z > 6$) and
%their observer-frame optical emission is absorbed by the IGM. The majority,
%however, suffer from large dust obscuration (e.g., Perley et al. 2009, AJ, 138,
%1690; Greiner et al.\ 2011, A\&A, 526, 30). Identifying such GRBs is important
%for constraining the fraction of obscured star formation. In both cases, NIR
%emission is expected. X-shooter can adequately study these objects, provided
%that a NIR counterpart is timely identified, for which we have the dedicated
%HAWKI run D.
%
%The detection of {\bf absorption line variability} can reveal the burst
%influence on the surrounding medium and in turn the absorber distance from the
%burst and its metallicity (Vreeswijk et al.\ 2007, A\&A 468, 83; D'Elia et al.\
%2009, ApJ, 694, 332; Th\"one et al.\ 2011, MNRAS, 414, 479; De Cia et al.\ 2012,
%A\&A, 545, 64; Hartoog et al.\ 2013). {\bf Short GRBs} originate in a
%substantially different environment compared to long GRBs. Short GRBs may be
%related to the merging of compact binaries and the coalescence time can be long
%enough to allow the progenitor system to move far away from the star formation
%site (Belczynski et al.\ 2002, ApJ, 571, 147). Up to now, however, no spectrum
%with a sufficient signal-to-noise ratio of a short GRB afterglow has been
%secured. A knowledge of the redshift distribution of short bursts is of key
%importance for the next generation of gravitational wave experiments, as they
%are the likely EM counterparts to their primary targets.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria and follow-up}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample selection criteria} \label{samplecrit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists already, being it weather conditions, pointing restrictions
of the telescope or poorly localized bursts as reported by the alerting
telescope. To maximize the return of the follow-up campaign we have chosen a few
selection criteria that attempts to provide an unbiased selection of bursts,
while allowing for a high follow-up success rate. The importance of defining
unbiasing selection criteria has been highlighted previously
\citep{Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a}, when trying to
address the specific underlying distribution functions such as the redshift
distribution, host metallicity distribution, or brightness distribution. When
investigating a specific distribution function, a high degree of
\textit{completeness} is desired meaning that in terms of the parameter of
interest, the largest possible fraction of targets in the sample has
determinations of that parameter \citep{Perley2016b}.

We simultaneously want to minimize any biases against astrophysical conditions
while at the same time maximizing likelihood of observations allowing us to
obtain a higher degree of observing completeness. By restricting the selection
criteria to conditions local to Galaxy and therefore irrespective of intrinsic
GRB properties, the hope is that the sample collect represents the underlying
distribution of GRBs in a fair way while at the same time, additional unbiasing
cuts can be imposed that can improve the follow-up success. The selection
criteria used here is based on previous successes \cite{Jakobsson2006b,
	Fynbo2009, Hjorth2012}, which has shown to represent the underlying distribution
to a high degree. We test the degree of completeness in Sect.
\ref{completeness}.

The selection criteria are defined as follows:

\begin{enumerate}
	\item GRB triggered onboard by Swift.
	\item XRT started observing within 10 minutes since the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12.
	\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. The below criteria are however those
	defining the sample.}
	\item No bright closeby stars.
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

Because our ability to observe GRB afterglows is strongly correlated with the
rapidness and precision of the target positions delivered by the triggering
facilities, by selecting only bursts that have been triggered on board the
\textit{Swift} space telescope \cite{Gehrels2004}, by the Burst Alert Telescope
(BAT) \citep{Barthelmy2005}, we start out with a homogeneously selected sample
where burst characteristics are delivered immediately, allowing for an informed
follow-up strategy. Despite the complexity of the triggering mechanism on board
\textit{Swift} \citep{Band2006}, attempts at inferring properties of the
underlying cosmic GRB population based on the detection thresholds and
triggering algorithms have been made \citep{Lien2014, Graff2016} and thereby by
restricting the follow-up effort to bursts detected by \textit{Swift}, the
limitations of the parent sample is well studied.

Because the localization accuracy of BAT is 1 - 4 arcminutes
\citep{Barthelmy2005}, we additionally require an X-ray position from the X-ray
Telescope (XRT) \citep{Burrows2005} to be distributed to the GCN network
\citep{Barthelmy2000} with 10 hours and to account for observing constraints on
\textit{Swift} that XRT began observations within 10 minutes. The additional
rapidness requirement of the XRT follow-up means that the number of detected
X-ray afterglows for BAT triggers is 100 percent \citep{Burrows2007}. The
\textit{Swift} constraints on the GRB follow-up means that the sample presented
here inherits the \textit{Swift} sample properties, except the additional XRT
constraint. Because the XRT completeness is very high\citep{Burrows2007}, this
should change the parent sample significantly. This issue is addressed again in
Sect. \ref{completeness}.

To allow for observability we additionally require that the GRB is visible
(duh!) from the telescope site at Cerro Paranal, Antofagasta, Región de
Antofagasta, Chile, for a least 1 hour allowing time for the spectroscopic
observations to completed. Since the GRB population is isotropically distributed
seen from Earth and that GRB properties does not depend on position on the sky
\citep{Briggs1996}, this cut does not influence our ability to sample the
underlying GRB population. The same arguments applies to the requirement that
there are no nearby bright stars, which is additionally supplemented by the
requirement what the galactic dust content does not exert an extinction higher
than $A_V \lesssim 0.5$ mag based on the extinction maps by
\citet{Schlegel1998}. Because of  isotropy, these additional cuts should not
influence the optical properties of the bursts themselves, only our ability to
successfully secure observations that allows us to investigate the spectroscopic
properties of GRBs.

We address the succesfullness of these criteria and the actual follow-up effort
in representing the full \textit{Swift} sample ion Sect. \ref{completeness}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up procedure and practicalities}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Will be written by Daniele?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Observations} \label{obs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}}
	\caption{Mollweide projection in galactic coordinates of the full sky showing
		the positions on the sky of the bursts presented in this work. The empty
		circles are the positions of all the \textit{Swift} bursts detected until the
		end of 2016. In blue is shown the position of the bursts fulfilling the sample
		criteria specified in Sect. \ref{samplecrit}. The red circles are added around
		GRBs which have been followed up by X-shooter since the commissioning of the
		instrument. Blue points with red circles represent GRBs that enter our sample
		and have X-shooter spectroscopy. The different samples are compared in Sect.
		\ref{results}. The background shows the dust maps presented in
		\citet{Schlegel1998} where the sample criteria cut with $A_V \lesssim 0.5$ mag
		is removed.}
	\label{fig:skymap}
\end{figure*}

Observations in this sample has been carried out using the cross-dispersed
echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on two of the four
Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during the duration
of this follow-up campaign, which covers the entire lifetime of X-shooter.
Observations have been carried out from the ESO period P84 through P98 under the
following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079, 095.A-0045,
094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088, 089.A-0067,
088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260. A few additional
bursts, not a part of the \textit{statistical sample}, have been included from
092.D-0056(A) (PI: Rau), 092.D-0633(E) (PI: Greiner), and Levan under
095.B-0811(B) (PI: Levan). This sample represents \textit{all} GRBs that have
ever been followed up by X-shooter.

The first GRB followed up is GRB090313, observed 15th of March 2009, which was
while X-shooter was mounted on UT3 during the commissioning of the instrument.
The bursts observed during the commissioning period is not a part of our
statistical sample, but are nonetheless published as part of the bursts observed
by the X-shooter GRB team. The first burst observed after science verification
was completed when X-shooter was moved to UT2 is GRB091018 which thereby
constitute the first burst entering our statistical sample. For all bursts that
fulfill our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up has been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of this
effect is included in Sect. \ref{badbursts}.

@@

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex} and plot the positions off all the bursts on
the night sky in Fig. \ref{fig:skymap}. The central zone of avoidance due to
high galactic extinction cutoff is visible with an apparent isotropic
distribution of the bursts away from the galactic center except in the upper
right corner, where Earth casts a shadow on the sky as seen from the telescope.

A fraction of the bursts presented here \todo{Insert exact
number} have already had their hosts investigated in \citet{Kruhler2015}, for
which extractions of the hosts exist. The focus of the data presented here are
on the afterglows themselves and in the absence of a clear afterglow, the host.
We will not, however, investigate the hosts.

@@

The biased introduced in the redshift distribution by the increased ability to
secure redshift of optically brighter bursts has been investigated by
\citet{Turpin2016} which find that we systematically find redshift for longer
GRBs. Additionally only the brightest GRBs are seen above redshift $z \gtrsim
1$. \todo{Additional text}

@@

\clearpage

>>>>

\usepackage{url}
\usepackage{color,hyperref}
\definecolor{linkcolor}{rgb}{0,0,0.7}
\hypersetup{colorlinks=true,
	linkcolor=linkcolor, 
	citecolor=linkcolor,
	filecolor=linkcolor, 
	urlcolor=linkcolor}

@@

\usepackage{xcolor}
\newcommand\todo[1]{\textcolor{red}{(#1)}}

@@

\title{\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample}
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}

\titlerunning{XSGRB sample}

\author{
	J.~Selsing\inst{1}, 
	T.~Kr\"{u}hler\inst{1}, 
	P.~Goldoni\inst{14}, 
	D.~Malesani\inst{1}, 
	J.~P.~U. Fynbo\inst{1}, 
	A.~de~Ugarte~Postigo\inst{11}, 
	J.~Japelj\inst{20},
	P.~D'Avanzo,
	Z.~Cano,
	S.~Covino\inst{10}, 
	V.~D'Elia\inst{7, 12}, 
	H.~Flores,
	O.~E.~Hartoog\inst{6},
	J.~Hjorth\inst{1}, 
	P.~Jakobsson\inst{5}, 
	A.~Levan,
	A.~Melandri,
	S.~Piranomonte \inst{7},
	R.~S\'anchez-Ram\'irez\inst{11},
	S.~Schulze\inst{17, 18}, 
	N.~R.~Tanvir\inst{19},
	C.~Th{\"o}ne,
	S.~D.~Vergani\inst{7, 8},
	P.~M.~Vreeswijk\inst{3}, 
	D.~J.~Watson\inst{1},
	K.~Wiersema\inst{19},
	D.~Xu\inst{1}
	L.~Christensen\inst{1},
	A.~De~Cia\inst{3}, 
	L.~Kaper\inst{6}, 
	L.~A.~Antonelli,
	F.~Fiore,
	A.~Gomboc,
	P.~Groot,
	F.~Hammer,
	C.~Ledoux\inst{2}, 
	E.~Maiorano,
	B.~Milvang-Jensen\inst{1}, 
	E.~Palazzi,
	E.~Pian,
	J.~Schaye,
	G.~Tagliaferri\inst{7},
	R.~A.~M.~J.~Wijers\inst{6}
}

@@

\abstract{
The {\it Swift} satellite allows us to use gamma-ray bursts (GRBs) to peer into 
the hearts of star forming galaxies through cosmic time. Our open
collaboration, representing most of the active ESO member researchers in this
field, seeks to build a public legacy sample of GRB X-shooter spectroscopy
while {\it Swift} continues to fly. We propose to continue our programme to target
all suitably observable GRB afterglows (up to 15 bursts per semester), with the
primary goal of producing a well-defined, homogeneous, statistically useful
sample. To date, our spectroscopy covers a redshift range from 0.059 to about
8, with more than 20 robust metallicity measurements from absorption lines (over 
the redshift range 1.7--5.9) and 4 secure detections of H$_2$ or CH molecular
absorption. Such information is extremely difficult to obtain by other means.
In terms of studying the spread and redshift evolution in gas-phase properties, the
sample is still limited by low-number statistics. 
}

@@

GRBs are interesting as they provide the opportunity to derive new insights
into a very wide range of topics in astrophysics. In that sense GRB
research is very multidisciplinary. The {\it Swift} satellite \citep{Gehrels09}, 
which was launched in 2004, has made
it possible to harvest the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by
results from earlier missions.

A crucial aspect of the great success of the {\it Swift} mission has been
the extensive ground based follow-up observations of the afterglows and of
the host galaxies of the GRBs. This work has involved a large community
of researchers both inside and outside the official {\it Swift} team. This fruitful
collaboration has been greatly facilitated by the open data access policy
of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs (typically
R=$\lambda/\Delta\lambda$$<$1000) \citep[e.g.][]{Fynbo2009}. For a handful of
very bright afterglows high-resolution (typically R$>$20000) spectra were
secured.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd generation
instruments at the ESO Very Large Telescope (VLT). It is a high-throughput 
echelle spectrograph
covering the full spectra range from the UV atmospheric cutoff to the near-infrared
(near-IR)
K-band with three spectral arms (UVB, VIS, NIR). It was designed very much with
transient source follow-up in mind as the fading luminosities of such sources 
makes in urgent to secure as extensive coverage as possible in the shortest possible 
time. At the same time the resolution was set to be in the range 4000--9000 in order
to be able to get a large useful spectra range between the many sky-background
emission lines in the red and near-IR.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria and observations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample selection criteria} \label{samplecrit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists from already, being it weather conditions, pointing
restrictions of the telescope or poorly localized bursts as reported by the
\textit{Swift}-telescope. To maximize the return of the follow-up campaign
we have chosen a few selection criteria that attempts to provide an unbiased
selection of bursts, while allowing for a high success-rate

\footnote{Note that in the P84 proposal the criteria have
been stated a bit differently, the visibility constraint being replaced by a
declination + Sun angle constraint. The above criteria are however those
defining the sample.}

\begin{enumerate}
	\item GRB triggered onboard by Swift.
	\item Galactic $A_V \lesssim 0.5$ mag.
	\item XRT started observing within 10 minutes since the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12.
	\item No bright closeby stars.

\end{enumerate}

A significant fraction of the bursts presented here \todo{Insert exact
number} have already had their hosts investigated in \citet{Kruhler2015}, for
which extractions of the hosts exist. The focus of the data presented here are
on the afterglows themselves and in the absence of a clear afterglow, the host.
We will not, however, investigate the hosts.

We simultaneously want to minimize any biases against astrophysical conditions
while at the same time maximizing likelihood of observations. By restricting
the selection criteria to conditions local to Galaxy.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up procedure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Observations} \label{obs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

All the observations in this sample has been carried out using the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
two of the four Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal)
during the duration of this follow-up campaign, which covers the entire lifetime
of X-shooter. Observations have been carried out from the ESO period P84
through P98 under the following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079,
095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260.

The first GRB followed up is GRB090313, observed 15th of March 2009, which was
while X-shooter was mounted on UT3 during the commissioning of the instrument.
The bursts observed during the commisioning period is not a part of our
statistical sample, but are nonetheless published as part of the bursts observed
by the X-shooter GRB team. The first burst observed after science verification
was completed when X-shooter was moved to UT2 is GRB091018 which thereby
constitute the first burst entering our statistical sample. For all bursts that
fulfill our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up has been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of this
effect is included in Sect. \ref{badbursts}.

@@

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex}

@@

@@

\newpage
\appendix
